Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-06-05 Manuscript/Manuscript.Rmd
MS: discussion, abstract edits, refs added

---
title: Nitric oxide feedback to ciliary photoreceptor cells orchestrates UV avoidance
  in zooplankton
author: 
  - name: "Kei Jokura<sup>1,`*`</sup>, Nobuo Ueda<sup>2</sup>, Martin Gühmann<sup>3</sup>, Luis Alfonso Yañez-Guerra<sup>1</sup>, Piotr Słowiński<sup>1</sup>, Kyle C. A. Wedgewood<sup>1</sup> and Gáspár Jékely<sup>1,4,`*`</sup>"
    affiliation: "<sup>1</sup>Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, United Kingdom <sup>2</sup>Okinawa Institute of Science and Technology, Okinawa, Japan, <sup>3</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom, <sup>4</sup>Centre for Organismal Studies (COS), University of Heidelberg, 69120 Heidelberg, Germany  <br> <sup>*</sup>Correspondence`:` k.jokura @ exeter.ac.uk, gaspar.jekely @ cos.uni-heidelberg.de"
output:
  html_document:
    theme: lumen
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key regulator of animal physiology.
Here we uncover a role for NO in shaping circuit dynamics to orchestrate light-avoidance behaviour.
We studied UV/violet light avoidance mediated by brain ciliary photoreceptors (cPRCs) in larval *Platynereis dumerilii*, a marine annelid.
We found NOS expressed in interneurons (INNOS) postsynaptic to cPRCs.
Stimulation of cPRCs by violet light leads to cPRC inhibition but concomitant INNOS activation and NO production.
NO feeds back to cPRCs and triggers their delayed activation through an unconventional guanylyl cyclase.
This results in the activation of projection interneurons and the inhibition of serotonergic ciliomotor neurons.
In *NOS* mutants, NO feedback and projection-neuron activation do not occur and avoidance behaviour is defective.
By mathematical modelling, we could recapitulate phototransduction and circuit dynamics in both wild-type and mutant larvae.
Our results reveal how NO-mediated retrograde signalling gates a synaptic circuit to initiate light-avoidance behaviour.

@@

In nervous systems, synaptic transmission and volume transmission together shape circuit dynamics.
While synaptic transmission occurs at specialised contact sites, volume transmission is characterised by the delocalised release of diverse diffusive neuromodulators.

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized by nitric oxide synthase (NOS) from L-arginine is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca2+/calmodulin-dependent activation of NOS following calcium increase, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC) cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

NO signalling has been extensively studied in the vertebrate retina where NOS is expressed in amacrine, ganglion as well as other cells [@Cudeiro1999; @Wang2007; @jacoby2018].
The actions of NO can be diverse and not always involve the canonical sGC-cGMP pathway [@Tooker2013; @Wei2012; @jacoby2018].
Defective NO signalling in NOS knockouts for example leads to a decreased sensitivity of retinal ganglion cells to light stimulation [@Wang2007].
However, due to the complex expression of NOS in vertebrates and the diversity of its functions, it has been challenging to link the neurophysiological effects of NO to signalling mechanisms and behaviour change.

@@

Due to these difficulties, we still know little at the whole organism level about how NO production relates to stimulus conditions, how it shapes circuit activity at specific neuron types, and how NO-dependent modulation relates to behaviour.

To investigate NO function in neural circuit dynamics and behaviour, here we study larvae of the marine annelid model *Platynereis dumerilii* [@ozpolat2021].
*Platynereis* has emerged as a systems neuroscience model where behaviour and activity imaging can be combined with genetic manipulations and a whole-body synaptic connectome and gene expression atlases are available [@Vergara2021; @Veraszto2017; @veraszto2020; @ozpolat2021].

@@

Retrograde signalling by NO from INNOS to cPRC leads to the sustained activation of cPRCs and postsynaptic neurons even after the end of stimulation.
This activated state is maintained for several tens of second.
NO signalling thus induces a transient circuit state or short-term memory trace in the *Platynereis* larval brain.
Because of the short life time of NO, this molecule may be well suited to encode transient memory traces [@Kuntz2017].
In the ellipsoid body of the *Drosophila* central brain, NO signalling has a similar function.
Here, NOS is specifically expressed in the R3 ring neurons and is required for the short-term (\~4 sec) visual memory of objects [@Kuntz2017].
NO is produced in the axons of R3 neurons and acts directly on sGC in the same axons.
This autocrine signal leads to a CNG-dependent temporary increase in calcium levels, carrying the working memory trace [@Kuntz2017].

@@

In *Drosophila*, NOS of PPL1-γ1pedc shortens memory retention while promoting fast memory updating in response to new experiences, indicating that NO regulates forgetting [@Aso2019].

@@

UV/violet light induces downward swimming head down [@Veraszto2018].
Since the direction of UV/violet light is not important for this, swimming direction must be determined by gravity.

Connectome reconstruction and whole-body cell annotation in the three-day-old *Platynereis* larva has not revealed any balancer organ to sense orientation in the gravity field [@bezares-calderon2019; @veraszto2020].
It follows that body orientation must be determined by physical parameters including the buoyancy, centre of mass and shape of the larva as well as differential ciliary activity.
Except for ciliary activity, all these parameters are likely invariant during the UV response. We found a UV-dependent inhibition of the Ser-h1 and activation of the MC cell consistent with the inhibition of ciliary beating [@veraszto2017].

Our current hypothesis about the mechanism of downward orientation posits that the reduced beating of prototroch cilia combined with the invariant beating of trunk cilia result in a head-down stable state due to physics alone. In contrast, during the pressure response prototroch cilia beat faster than trunk cilia, leading to a head-up orientation [@bezares-calderon2023]. Testing this hypothesis will require biophysical experiments and modelling.

## UV-avoidance circuits

It is known that complex integrative networks of neurons are involved in UV avoidance behaviour in planaria [@Shettigar2021]

>>>>

---
title: Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit
author: 
  - name: "Kei Jokura<sup>1,`*`</sup>, Nobuo Ueda<sup>2</sup>, Martin Gühmann<sup>3</sup>, Luis Alfonso Yañez-Guerra<sup>1</sup>, Piotr Słowiński<sup>1</sup>, Kyle C. A. Wedgewood<sup>1</sup> and Gáspár Jékely<sup>1,4,`*`</sup>"
    affiliation: "<sup>1</sup>Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, United Kingdom <sup>2</sup>Okinawa Institute of Science and Technology, Okinawa, Japan, <sup>3</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom, <sup>4</sup>Centre for Organismal Studies (COS), University of Heidelberg, 69120 Heidelberg, Germany  <br> <sup>*</sup>Correspondence`:` k.jokura @ exeter.ac.uk, gaspar.jekely @ cos.uni-heidelberg.de"
output:
  html_document:
    theme: lumen
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key regulator of animal physiology.
Here we uncover a role for NO in gating a light-avoidance circuit and mediating short-term memory of UV exposure.
We studied UV/violet light avoidance mediated by brain ciliary photoreceptors (cPRCs) in larval *Platynereis dumerilii*, a marine annelid.
We found NOS expressed in interneurons (INNOS) postsynaptic to cPRCs.
UV stimulation of cPRCs leads to INNOS activation and NO production.
NO signals retrogradely to trigger the sustained post-stimulus activation of cPRCs through an unconventional guanylyl cyclase.
This late activation inhibits serotonergic ciliomotor neurons via projection interneurons.
In *NOS* mutants, retrograde signalling, circuit output and UV-avoidance behaviour are defective.
By mathematical modelling, we recapitulate phototransduction and circuit dynamics in both wild-type and mutant larvae.
Our results reveal how NO-mediated retrograde signalling gates a synaptic circuit and induces short-term memory of UV exposure to orchestrate light-avoidance behaviour.

@@

In nervous systems, synaptic transmission and volume transmission together shape circuit dynamics [@bargmann2013].
While synaptic transmission occurs at specialised contact sites, volume transmission is characterised by the delocalised release of diverse diffusive neuromodulators.

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized by nitric oxide synthase (NOS) from L-arginine is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca2+/calmodulin-dependent activation of NOS following calcium increase, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC).
This can occur cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

In the vertebrate retina, NOS is expressed in amacrine, ganglion and other cells and the actions of NO can be diverse [@Cudeiro1999; @Wang2007; @jacoby2018].
For example, defective NO signalling in NOS knockouts leads to a decreased sensitivity of retinal ganglion cells to light stimulation [@Wang2007].
In the retina, NO signalling can also involve pathways other than canonical sGC signalling [@Tooker2013; @Wei2012; @jacoby2018].
Due to the complex expression of NOS in vertebrates and the diversity of its functions, it has been challenging to link the neurophysiological effects of NO to signalling mechanisms and behaviour change.

In the *Drosophila* brain, NO signalling is also involved in tuning circuit activity at diverse and specific sites.
In the ellipsoid body of the central complex, NO is involved in visual working memory [@Kuntz2017].
NO signalling also tunes the dynamics of associative memories in mushroom body circuits. NOS is expressed in PPL1-γ1pedc mushroom-body neurons where it is involved in shortening memory retention while promoting fast memory updating in response to new experiences [@Aso2019].

@@

Overall, we still know little about how NO production relates to stimulus conditions, how it shapes circuit activity at specific neuron types, and how NO-dependent modulation relates to behaviour.

To investigate NO function in neural circuit dynamics and behaviour, here we study larvae of the marine annelid model *Platynereis dumerilii* [@ozpolat2021].
*Platynereis* has emerged as a model for systems neuroscience where behaviour and activity imaging can be combined with genetic manipulations.
A whole-body synaptic connectome and gene expression atlases are also available [@Vergara2021; @Veraszto2017; @veraszto2020; @ozpolat2021].

@@

Retrograde signalling by NO from INNOS to cPRC leads to the sustained activation of cPRCs and postsynaptic neurons even after the end of stimulation.
This activated state is maintained for several tens of second.
NO signalling thus induces a transient circuit state or short-term memory trace in the *Platynereis* larval brain.
Because of the short life time of NO, this molecule may be well suited to encode transient memory traces [@Kuntz2017].

In the ellipsoid body of the *Drosophila* central brain, NO signalling has a similar function.
Here, NOS is specifically expressed in the R3 ring neurons and is required for the short-term (\~4 sec) visual memory of objects [@Kuntz2017].
NO is produced in the axons of R3 neurons and acts directly on sGC in the same axons.
This autocrine signal leads to a CNG-dependent temporary increase in calcium levels, carrying the working memory trace [@Kuntz2017].

@@

@@

UV/violet light induces downward swimming head down [@Veraszto2018].
Since stimulus direction is not relevant, swimming direction must be determined by gravity.

Connectome reconstruction and whole-body cell annotation in the three-day-old *Platynereis* larva has not revealed any balancer organ to sense orientation in the gravity field [@bezares-calderon2019; @veraszto2020].
It follows that body orientation must be determined by physical parameters including the buoyancy, centre of mass and shape of the larva as well as differential ciliary activity.
Except for ciliary activity, all these parameters are likely invariant during the UV response.
Our data suggest the UV-dependent inhibition of prototroch ciliary beating via Ser-h1 inhibition and MC activation [@veraszto2017].

We hypothesise that the differential beating of prototroch versus trunk cilia causes a head-up or head-down orientation due to physics alone.
During the UV response, prototroch cilia beat slower than trunk cilia, resulting in a head-down stable state ('rear-wheel drive').
In contrast, during the pressure response prototroch cilia beat faster than trunk cilia [@bezares-calderon2023], leading to a head-up orientation ('front-wheel drive').
Testing this hypothesis will require biophysical experiments and mathematical modelling.

## UV-avoidance circuits of extraocular photoreceptors

Animals evolved distinct photosensory systems coupled to non-overlapping circuits and guiding unique behavioural responses.
These sensory systems can employ different opsin molecules and be tuned to different wavelength of light.
The avoidance of noxious UV/violet light is often mediated by extraocular photoreceptors and their circuits, distinct from the pigmented visual eyes.
These two types of systems co-exist in *Platynereis* where the pigmented eyes and eyespots guide phototaxis behaviour with a maximum sensitivity to cyan light (\~500 nm) [@Randel2014a; @Veraszto2018].
Planarians also have pigmented visual eyes mediating phototaxis to cyan light and peripheral extraocular photoreceptors mediating UV avoidance behaviour [@Shettigar2021].
*Drosophila* larvae have cerebral eyes called Bolwig's organs involved in phototaxis [@kane2013] and several types of extraocular UV-sensory cells that tile the body wall [@Imambocus2022].

One common feature of UV-avoidance circuits is their sustained post-stimulus activation following UV exposure.
This can involve peptidergic signals as in planaria and maggots [@bray2023; @Imambocus2022] or NO as in *Platynereis*.
Volume transmission is well suited to integrate light exposure and maintain an internal state following noxious stimulation.
The amount of modulator released can scale with stimulus intensity or duration and maintain an altered circuit state due to the slower decay of the diffusive signals relative to synaptic transmission.
